---
title: "October 4 Lecture"
author: "Steve Hoffman"
date: "2022-10-04"
output: 
  ioslides_presentation:
  widescreen: TRUE
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(
	echo = TRUE
)
```

## LOAD PACKAGES 
```{r}
library(tidyverse)

# James used "moments" for the skewness calculations
# Nice, James!
library(moments)
DAM1.data <- read_csv(file = "DAM1.csv") 
```

## Nice work, Josephine
```{r}
#clean the dataset
DAM1.clean <- DAM1.data %>%
  mutate(Class = factor(Class, levels=c("1", "2", "3"),
                        labels=c("Ms. Atwell", "Ms. Barnes","Mr.Campbell")),
         Gender = ifelse(is.na(Gender), "nonbinary", Gender),
         Gender = factor(Gender, levels=c("1","2","nonbinary"),
                         labels=c("boys","girls","nonbinary")))

DAM1.clean

```

---
```{r}
print(DAM1.clean, n = 54)
```

## Section 3: Spelling Score Boxplot
and get rid of the y axis labels
```{r spelling box, echo=FALSE}
# Boxplot for spelling
ggplot(data = DAM1.clean, mapping = aes(x = Spelling)) +
  geom_boxplot() +
  labs(title = "Boxplot of Spelling Scores for All Students") +
  xlab("Spelling Test Score") +
  scale_y_continuous(labels = NULL) + # 
  theme_classic()
```

## Look at the quantiles
```{r}
quantile(DAM1.clean$Spelling, c(0.25, 0.5, 0.75))
62-48.25 ## Look at you using R as a calculator ;)
```



## Section 9: Spelling Score Boxplots by Classroom
```{r, echo=TRUE}
# Boxplots of spelling for each of the three classrooms
ggplot(data = DAM1.clean, mapping = aes(x = Spelling, y = Class )) + 
  geom_boxplot() +
  labs(title = "Boxplots of Spelling Scores by Classroom Teacher") +
  xlab("Spelling Test Score") + 
  theme_classic() 
```

---

```{r, echo=TRUE}

DAM1.clean %>% 
  group_by(Class) %>%
  summarize(first=quantile(Spelling,probs=0.25),
            second=quantile(Spelling,probs=0.5),
            third=quantile(Spelling,probs=0.75))
  62.8-55
  59.8-48.2
  62-46
```

## What Hoffman dreams he will see next DAM

I learned a lot by assigning this project to you. And I very much appreciate the effort. This isn’t easy stuff — even the basic aspects of writing descriptively about univariate statistics take work. 

## Debate each other when writing

We don’t know the kids from this assignment. And one thing that I was hoping that you would do is debate what you might say about any number of issues. For example, we don’t know which students have an IEP (let’s say it could be Charles and Nathan, who would be within Ms Barnes’s boxplot, but are outliers because Ms Atwell’s class mainly full of great spellers). We weren’t asked to comment on that, but keep in mind that virtually every school serves students with learning disabilities and other needs. And it might be that one of the teachers is assigned several of the students with the most needs because the Principal knows just how skilled they are.

## Is this fake dataset realistic?
And now we are looking at data and can think about what it SHOULD look like. For example, Mary and Gray noted that perhaps the math test was harder than the spelling test. That’s probably true. 

## Send me .Rmd files

What I want is to make your .Rmd files easily loaded on my computer. You can probably even pass them back and forth. (Someday you’ll be working on GitHub but that’s not this term)

So, right now, before we go on:

Open a new .Rmd file on your computer. Write me a quick note in place of the stuff that the RMarkdown people provide as starters.

And send it to me in the assignment called "Practice .Rmd Assignment"

(I might show your assignment to the class, so make sure you don't send anything embarrassing...)


## Last questions about DAM1

After this, we're going to switch to some updated Google slides. 

And then we're going to look at the NAEP website together.

And then we'll pop over to this slide deck for some more ggplot pictures and stuff that you might find helpful.

Also, side note: You have a quiz later this week. It’s going to be open-book, open notes (but closed to working together). And you’ll only have one shot, not several, to answer the MC items. 

## LTT Math Age 9

```{r}
library(tidyverse)
library(readxl) 

# Read Excel file
dat <- read_excel("LTT_Math9__1978_2022.xlsx")
```

## Clean dataset
```{r}
#as tibble
dat <- as_tibble(dat)

# make Format a factor
dat <- dat %>%
  mutate(Format = factor(Format, levels=c("NotOriginal", "OriginalFormat"), labels=c("Not Original Format", "Original Format")))
```

## The data
```{r}
print(dat, n = 34)
```


## Add columns calculating difference in math scores since 1978
Long-term Trend national average scores in 1978:
Average Math LTT in 1978 Age 9: 219
SD of math9 scores in 1978: 36

```{r}
dat <- dat %>%
  mutate(Diff78_9M = as.integer(Math9 - 219))

# Add column calculating difference in SD since 1978
dat <- dat %>%
  mutate(Diff78sd_9M = Diff78_9M/SD9)
```

## Create plot 
```{r}
# Create plot for 9-year-olds similar to page 94 in Measuring Up
LTTmath9.plot <- ggplot(dat) +
  geom_line(aes(Year, Diff78sd_9M, group = Format)) +
  theme_classic() +
  labs(x = "Year",
       y = "Difference in Standard Deviation") +
  labs(title = "LTT National Average Math Score 9-year-olds", 
       subtitle = "Difference in average scores (in SD) since 1978") +
  scale_x_continuous(limits = c(1978, 2022))
```

---

```{r}
LTTmath9.plot   
```

    
## Create plot by gender
```{r}
dat <- dat %>%
  mutate(Diff78_math_female = FemaleAvg - 219) %>%
  mutate(Diff78_math_male = MaleAvg - 219) %>%
  mutate(Diff78sd_math_female = Diff78_math_female/FemaleSD) %>%
  mutate(Diff78sd_math_male = Diff78_math_male/MaleSD)

LTTmath9_gender.plot <- ggplot(dat) +
  geom_line(aes(Year, Diff78sd_math_female, group = Format, color = "Girls")) +
  geom_line(aes(Year, Diff78sd_math_male, group = Format, color = "Boys")) +
  theme_classic() +
  labs(x = "Year",
       y = "Difference in Standard Deviation") +
  labs(title = "LTT National Average Math Score 9-year-olds", 
       subtitle = "Differences by gender since 1978") +
  scale_x_continuous(limits = c(1978, 2022))
```

---

```{r}
LTTmath9_gender.plot   
```


## Create plot by race

```{r}
dat <- dat %>%
  mutate(Diff78_math_white = WhiteAvg - 219) %>%
  mutate(Diff78_math_black = BlackAvg - 219) %>%
  mutate(Diff78_math_hispanic = HispanicAvg - 219) %>%
  mutate(Diff78_math_asian = AsianAvg - 219) %>%
  mutate(Diff78sd_math_white = Diff78_math_white/WhiteSD) %>%
  mutate(Diff78sd_math_black = Diff78_math_black/BlackSD) %>%
  mutate(Diff78sd_math_hispanic = Diff78_math_hispanic/HispanicSD) %>%
  mutate(Diff78sd_math_asian = Diff78_math_asian/AsianSD) 

LTTmath9_race.plot <- ggplot(dat) +
  geom_line(aes(Year, Diff78sd_math_white, group = Format, color = "White")) +
  geom_line(aes(Year, Diff78sd_math_black, group = Format, color = "Black")) +
  geom_line(aes(Year, Diff78sd_math_hispanic, group = Format, color = "Hispanic")) +
  geom_line(aes(Year, Diff78sd_math_asian, group = Format, color = "Asian")) +
  theme_classic() +
  labs(x = "Year",
       y = "Difference in Standard Deviation") +
  labs(title = "LTT National Average Math Score 9-year-olds", 
       subtitle = "Differences, by race, since 1978") +
  scale_x_continuous(limits = c(1978, 2022))
```

---

```{r}
LTTmath9_race.plot    
```


## Normal Curves 2022 LTT Math

```{r}
normal_plots <- ggplot(data = data.frame(x = c(100, 360)), aes(x)) +
  stat_function(fun = dnorm, n = 590, args = list(mean = 244, sd = 37), color = "Pink") + 
  stat_function(fun = dnorm, n = 140, args = list(mean = 212, sd = 42), color = "Black") +
  stat_function(fun = dnorm, n = 190, args = list(mean = 223, sd = 40), color = "Brown") + 
  stat_function(fun = dnorm, n = 60, args = list(mean = 257, sd = 39), color = "Dark Green") +
  scale_y_continuous(breaks = NULL) +
  ylab("") +
  ggtitle("LTT estimated math scores for 9-year-olds in 2022")
```

---

```{r}
normal_plots
```




